---
title: 'Polysomes profiling analysis'
author: "Cécile Lecampion"
date: "`r format(Sys.time(), '%B, %Y')`"
output:
  html_document:
    theme: cerulean
    highlight: tango
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
  pdf_document:
    toc: yes
    toc_float: true
    number_sections: yes
---
***
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

Data of polysome profiling are collected from spectrophomtometer. 
Results are `.csv` file and are organized like this :

3 columns looking like : 

0.0001041666692,0.1534467638,<br>
0.000312499993,0.1535461843,<br>
0.0005208333605,0.1536204666,<br>
...

Only the second column is kept. There is some text in the file. Only numeric values are kept.
Two first lines and useless lines at the end of the file are ommitted

```{r, message=FALSE, echo=FALSE, warning= FALSE}
# Function to load the data
# Load the csv file and return a one cilumon data frame
# Column is named with parameter colName
# Usage exemple :
#    df <- f_load_polysome_csv("a1.csv", "a1")
#__________________________________________________________________________________________________
f_load_polysome_csv <- function(csvFile, colName) {
  lines <- readLines(csvFile)
  linesNbToImport <- grep("^\\d+\\.\\d+,\\d+\\.\\d+,", lines, perl = TRUE)
  df <- read.table(text = lines[linesNbToImport], header = FALSE, sep = ",", dec = ".")
  df <- as.data.frame(df$V2)
  colnames(df) <- c(colName)
  return(df)
}
#__________________________________________________________________________________________________
```

```{r, message=FALSE, echo=FALSE, warning= FALSE}
# THIS CHUNK HAS TO BE MODIFIED FOR EACH ANALYSIS
# Loading data
# Give data a generic name to simplify the command in the analysis
a1 <- f_load_polysome_csv("line1-1.csv", "a1")
a2 <- f_load_polysome_csv("line1-2.csv", "a2")
b1 <- f_load_polysome_csv("line2-1.csv", "b1")
b2 <- f_load_polysome_csv("lien2_2.csv", "b2")
c1 <- f_load_polysome_csv("line3-1.csv", "c1")
c2 <- f_load_polysome_csv("line3-2.csv", "c2")

# Keep the generic names of the data in a variable
NAMES <- c("a1", "a2", "b1", "b2", "c1", "c2")

if (!require(dplyr)) { install.packages('dplyr')}
library(dplyr)

if (!require(purrr)) { install.packages('purrr')}
library(purrr)

# Create a liste and build a data frame with all data
liste <- list(a1, a2, b1, b2, c1, c2)
df <- liste %>% map( ~ .x %>% tibble::rownames_to_column('rn')) %>% reduce(full_join, by = 'rn')
```

# Varables

Define the variable for the analysis

```{r, message=FALSE, echo=FALSE, warning= FALSE}
# Limits for graphs axis
# On X axis
XMIN <- 1000
XMAX <- 3600

# On Y axis
YMIN <- 0
YMAX <- max(df[XMIN:XMAX, 2: ncol(df)], na.rm = TRUE) + (max(df[XMIN:XMAX, 2: ncol(df)], na.rm = TRUE))*0.1

# Area where to search for the monosome peak
ZMONO <- c(2200 : 3000)

# Area for the valley
# Area before the monosome
Z_BMONO <- c(1500:2000)
# Area after the monosome
Z_PMONO <- c(2100:2600)
# Area after the peak in supernatent
Z_PSUR <- c(2800:3200)

# Limits to compute replicats mean
MEAN_START <- 500
MEAN_END <- 5000

# Real names of the data
# Keep here the real names of the data
DATA_REAL_NAME <- c("Line1", "Line2", "Line3")

# Number of replicats
NB_REPLICAT <- 2
```

**Limits for graphs axis**<br>  

* On X axis :<br>
`r {"XMIN: "}`
`r XMIN`<br>
`r {"XMAX: "}`
`r XMAX`

* On Y axis :<br>
`r {"YMIN: "}`
`r YMIN`<br>
`r {"YMAX: "}`
`r YMAX`

**Area where to search for the monosome peak**<br>
`r {"ZMONO: "}`
`r c(ZMONO[1], ZMONO[length(ZMONO)])`

**Area for the valley**<br>  

* Area before the monosome<br>
`r {"Z_BMONO: "}`
`r c(Z_BMONO[1], Z_BMONO[length(Z_BMONO)])`

* Area after the monosome<br>
`r {"Z_PMONO: "}`
`r c(Z_PMONO[1], Z_PMONO[length(Z_PMONO)])`

* Area after the peak in supernatent<br>
`r {"Z_PSUR: "}`
`r c(Z_PSUR[1], Z_PSUR[length(Z_PSUR)])`

**Limits to compute replicats mean**<br>
`r {"MEAN_START: "}`
`r MEAN_START`

**Real names of the data**<br>
Keep here the real names of the data that will be used in the final figure<br>
`r {"DATA_REAL_NAME: "}`
`r DATA_REAL_NAME`

**Number of replicats**<br>
`r {"NB_REPLICAT: "}`
`r NB_REPLICAT`

# Data processing : monosome alignment
## Plot the raw data

```{r, message=FALSE, echo=FALSE, warning= FALSE, fig.cap="fig.1 : Raw polysomes data"}
par(mar=c(2,2,1,1))
plot.ts(df[XMIN:XMAX,2:ncol(df)],plot.type=c("single"), type = "l", 
        xlab = "", ylab = "",
        col=rainbow(ncol(df)))
```

## Identify the position of monosome peak
```{r, message=FALSE, echo=FALSE, warning= FALSE}
# Find the max between ZMONO and XMAX for each curves
if (!require(splus2R)) { install.packages('splus2R')}
library(splus2R)

mono <-(apply((apply(df[ZMONO, 2:ncol(df)], 2, peaks, span=201, strict=TRUE, endbehavior=0)), 2, which))+2199
# Index of the monosme peak is the minimum of all monosome peak
mono_pic <- min(mono)
```
`r {"The monosome peak is : "}`
`r mono_pic`

## Align all curves on monosomes
To align all monosome peak one have to remove the n first lines of each column. n is the difference between the monosome peak for the considered curve and the monosome peak as define by varaible mono_pic

```{r, message=FALSE, echo=FALSE, warning= FALSE}
# Compute the number of line to remove for each column
N <- mono-mono_pic

#__________________________________________________________________________________________________
# Function that remove the n first lines
remove_n_first_rows <- function(dat, n) if(n <= 0) dat else tail(dat, -n)
#__________________________________________________________________________________________________

# Remove the lines
Liste_mono_align <- mapply(remove_n_first_rows, df[, -1], N)

# Convert the list element to a data frame
Liste_mono_align <- lapply(Liste_mono_align, as.data.frame)
df_mono_align <- Liste_mono_align %>% map( ~ .x %>% tibble::rownames_to_column('rn')) %>% reduce(full_join, by = 'rn')
colnames(df_mono_align) <- c("rn",NAMES)

```

Plot the data to check that all monosmes peak are aligned

```{r, message=FALSE, echo=FALSE, warning= FALSE, fig.cap="fig.2 : Polysome profiles aligned on monosomes peak"}
par(mar=c(2,2,1,1))
plot.ts(df_mono_align[XMIN:XMAX,2:ncol(df_mono_align)],plot.type=c("single"), type = "l", 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_mono_align)))
```

# Data processing : Minimum of the curves to 0
Identify the minimum for each curve around the monosome and substract this value to each column

```{r, message=FALSE, echo=FALSE, warning= FALSE}
#Find the y_min around the monosome
y_min <-apply(df_mono_align[XMIN:mono_pic, -1], 2, min)

# Substract y_min 
df_zero_align <- as.data.frame(mapply('-', df_mono_align[, -1], y_min))

```

Plot the data to check that the minimum of each curve is now 0

```{r, message=FALSE, echo=FALSE, warning= FALSE, fig.cap="fig.3 : Polysome profiles aligned on 0"}
par(mar=c(2,2,1,1))
plot.ts(df_zero_align[XMIN:XMAX,2:ncol(df_zero_align)],plot.type=c("single"), type = "l", 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_zero_align)))
```

# Data processing : Mean of replicats and smoothing
## Mean of replicats

The mean is computed in row for the n column corresponding to the number of replicats (NB_REPLICAT)

```{r,  message=FALSE, echo=FALSE, warning= FALSE}
# row mean of n column with n the number of replicats: NB_REPLICAT 

# Define the number of lines/conditions in the analysis
nb_line <- ncol(df_zero_align)/NB_REPLICAT

# Select the column to porcess
select_col <- c(1:NB_REPLICAT)

# Create an empty data frame (length of the data frame is length of the result that will be storein it)
df_mean <- data.frame(matrix(ncol = 0, nrow = length(c(MEAN_START : MEAN_END))))

#__________________________________________________________________________________________________
# Function that return the row mean of the selected columns  
f_compute_rep_mean <- function (select_col) {
  result <- as.data.frame(rowMeans(df_zero_align[MEAN_START : MEAN_END, select_col]))
  return(cbind(df_mean, result))
}
#__________________________________________________________________________________________________

# Loop over the data frame
for (LineCount in c(1:nb_line)) {
  df_mean <- f_compute_rep_mean(select_col)
  select_col <- select_col+NB_REPLICAT
}
colnames(df_mean) <- DATA_REAL_NAME
```

Plot the data

```{r, message=FALSE, echo=FALSE, warning= FALSE, fig.cap="fig.4 : Polysome profiles mean of replicats"}
par(mar=c(2,2,1,1))
plot.ts(df_mean[XMIN:XMAX, ],plot.type=c("single"), type = "l", 
        ylim = c(0, YMAX-(min(y_min))), 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_mean)))
```

## smoothing with adjacent averaging method
```{r, message=FALSE, echo=FALSE, warning= FALSE}
if (!requireNamespace("forecast", quietly = TRUE)) { install.packages("forecast", repos = "http://cran.us.r-project.org") }
library(forecast)

df_smoothed <- as.data.frame(apply(df_mean, 2, ma, order = 50, centre = TRUE))
```

plot the data

```{r,  message=FALSE, echo=FALSE, warning= FALSE,results = FALSE, fig.cap="fig.5 : Polysome profiles"}
par(mar=c(2,2,1,1))
plot.ts(df_smoothed[XMIN:XMAX, ],plot.type=c("single"), type = "l", 
        ylim = c(0, YMAX-(min(y_min))), 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_smoothed)))
legend("topleft", legend = DATA_REAL_NAME,
       col = rainbow(ncol(df_smoothed)), lty = 1, lwd = 2,
       box.lty = 0)

# Save the plot
png("polysomes_profiles.png")

plot.ts(df_smoothed[XMIN:XMAX, ],plot.type=c("single"), type = "l", 
        ylim = c(0, YMAX-(min(y_min))), 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_smoothed)))
legend("topleft", legend = DATA_REAL_NAME,
       col = rainbow(ncol(df_smoothed)), lty = 1, lwd = 2,
       box.lty = 0)

dev.off()
```


# Data processing : compute area under the curve
## Check that the intervalls to search for the valley were correctly defined
```{r, message=FALSE, echo=FALSE, warning= FALSE, fig.cap="fig.6 : Checking the area for valley search"}
par(mar=c(2,2,1,1))
plot.ts(df_smoothed[, ],plot.type=c("single"), type = "l", 
        ylim = c(0, YMAX-(min(y_min))), 
        xlab = "", ylab = "",
        col=rainbow(ncol(df_smoothed)))
abline(v=c(Z_BMONO[1], Z_BMONO[length(Z_BMONO)]))
abline(v=c(Z_PMONO[1], Z_PMONO[length(Z_PMONO)]), col= "lightblue")
abline(v=c(Z_PSUR[1], Z_PSUR[length(Z_PSUR)]), col= "grey")
```

If the valley are contained in the intervals then the further analysis will be correct. If not, the intervals must be redefined at the begining and the anlysis must be started again.

## Compute area under the curve

```{r, message=FALSE, echo=FALSE, warning= FALSE}
# There is a shift of 1000 in the index value. wich.min() returns indexes that start at the first boundary of the search interval. 
# For real index add (boundary - 1) 
# To identifi index on the smoothed graph, remove 1000.

# Find min before monosme
min_poly <-apply(df_smoothed[Z_BMONO, ], 2, which.min)+1499

# find min after monosome
min_post_mono <-apply(df_smoothed[Z_PMONO, ], 2, which.min)+2099

# Find min after the peak in supernatent
min_post_sur <-apply(df_smoothed[Z_PSUR, ], 2, which.min)+ 2799
```

```{r, message=FALSE, echo=FALSE, warning= FALSE}

#Compute area under the curve
# add a column for indexes
df_smoothed$X <- c(1:nrow(df_smoothed))

# select the areas
# for polysome area
# create an empty list to store the results
List_auc_poly <- list()

for (i in c(1:nb_line)) {
  index1 <- XMIN
  index2 <- min_poly[[i]]
  df_auc_poly <- dplyr::filter(df_smoothed, X >= index1 & X <= index2) %>% 
    select(all_of(i), 4)
  List_auc_poly[[length(List_auc_poly)+1]]<- df_auc_poly
}

# for monosome area
# create an empty list to store the results
List_auc_mono <- list()

for (i in c(1:nb_line)) {
  index1 <- min_poly[[i]]
  index2 <- min_post_mono[[i]]
  df_auc_mono <- dplyr::filter(df_smoothed, X >= index1 & X <= index2) %>% 
    select(all_of(i), 4)
  List_auc_mono[[length(List_auc_mono)+1]]<- df_auc_mono
}

# for supernatent area
# create an empty list to store the results
List_auc_sur <- list()

for (i in c(1:nb_line)) {
  index1 <- min_post_mono[[i]]
  index2 <- min_post_sur[[i]]
  df_auc_sur <- dplyr::filter(df_smoothed, X >= index1 & X <= index2) %>% select(all_of(i), 4)
  List_auc_sur[[length(List_auc_sur)+1]]<- df_auc_sur
}

# Combine 3 areas in a data frame
df_AUC <- data.frame((matrix(ncol = length(DATA_REAL_NAME), nrow = 3)), 
                     row.names = c("AUC_poly", "AUC_mono", "AUC_sur"))
colnames(df_AUC) <- DATA_REAL_NAME

# load necessary packages
if (!requireNamespace("DescTools", quietly = TRUE)) { install.packages("DescTools") }
library(DescTools)

# Compute AUC
for (element in c(1:nb_line)) {
  AUC_poly <- AUC(x = List_auc_poly[[element]][,2], y = List_auc_poly[[element]][,1], na.rm = TRUE)
  df_AUC[1, element] <- AUC_poly
}

for (element in c(1:nb_line)) {
  AUC_mono <- AUC(x = List_auc_mono[[element]][,2], y = List_auc_mono[[element]][,1], na.rm = TRUE)
  df_AUC[2, element] <- AUC_mono
}

for (element in c(1:nb_line)) {
  AUC_sur <- AUC(x = List_auc_sur[[element]][,2], y = List_auc_sur[[element]][,1], na.rm = TRUE)
  df_AUC[3, element] <- AUC_sur
}


# Compute percent for further plot
for (line in c(1:nb_line)) {
  nom <- sprintf("%s%d ", "percent Line", line)
  df_AUC[,nom] <- 100*(df_AUC[,line]/sum(df_AUC[,line]))
}

```

Plot the results
<br>
<br>
```{r, message=FALSE, echo=FALSE, warning= FALSE, results = FALSE, fig.cap="fig.7 : Area under the curve"}

par(mar=c(2,2,1,1))
barplot(as.matrix(df_AUC)[, (nb_line+1) : ncol(df_AUC)], legend = c("Poly", "mono", "surpernatent"), 
        col = c("peachpuff", "thistle1", "lightsteelblue1"), 
        names.arg = DATA_REAL_NAME, width = c(40,40), xlim = c(0,240))

# Save the plot
png("area_under_curve.png")

barplot(as.matrix(df_AUC)[, (nb_line+1) : ncol(df_AUC)], legend = c("Poly", "mono", "surpernatent"), 
        col = c("peachpuff", "thistle1", "lightsteelblue1"), 
        names.arg = DATA_REAL_NAME, width = c(40,40), xlim = c(0,240))

dev.off()
```

# R packages used
Package	| Version |	Citation
--------|---------|---------
base	| 4.2.0	| R Core Team (2022)
DescTools |	0.99.44 |	Andri et mult. al. (2021)
forecast |	8.16 |	R. J. Hyndman and Khandakar (2008); R. Hyndman et al. (2022)
grateful |	0.1.11 |	Rodríguez-Sánchez, Jackson, and Hutchins (2022)
knitr |	1.39 |	Xie (2014); Xie (2015); Xie (2022)
rmarkdown |	2.14 |	Xie, Allaire, and Grolemund (2018); Xie, Dervieux, and Riederer (2020); Allaire et al. (2022)
splus2R |	1.3.3 |	Constantine and Hesterberg (2021)
tidyverse |	1.3.1 |	Wickham et al. (2019)


